### 2020-09-09 Manuscript/Response.to.reviewers.tex

Done.

\item{Line 99: The value of predicting which cluster (?community?) a given set of environmental parameters will fall within is not entirely clear to me. Is there a consistent set of taxa that make of up these clusters across time and space? Can specific groups of taxa be deduced from the ?community? number? Do the environmental variables accurately predict which set of taxa will be present? This section needs a bit more clarity.}

\textbf{Response:}

In sum, yes: these clusters are coherent groups of species and predictably occur in different environmental conditions. We have revised the main text to be more clear in this regard. {\color{red} RYAN Maybe add that some of the taxa are present in many clusters, and their differ in their relative proportin }

@@

We think the relationship between clusters and the environmental variables is important in a few aspects. First, as a means of ground truthing our findings, it has great value as the clusters were detected blind to the environmental data, and the CAP analysis return the distinctiveness of these clusters in that space - Figure 2B highlights the translation of those modified axis of the CAP plot into the real values of temperature and pH. Second, these clusters (maybe communities is a more committed term that requires a longer dataset in which to confirm co-ocurrence of members) do indeed reflect changes in the relative importance of the species present.

{\color{red} RAMON -- if we should use `cluster' instead of `community' we will need to change that in the main text. RYAN: Now that we think more carefully about it, it should be clusters of samples - so you use it when you talk about samples - and community of organism: which I agree should only be applied when you have  a deep understanding of the species composition.}

@@

\item{Line 102-104: According to my interpretation of Fig. 2C, it seems like the colder water communities actually had *lower* values of the given taxa, as the differences are all negative relative to other communities. It may help to specify in the text which characteristics are specific to which communities (communities 2 and 3 both span the lower temperature area of Fig. 2B).}

{\color{red} RGS: Figure 2C includes now taxa with the higher contribution to the dissimilarity between clusters, highlighting both positive and negative contributions.  }

@@

\item{Line 121: Refer to the relevant section of Figure 3 (A, B, or C)}

Done.

\item{Line 121-133: When referring to specific taxa as winners or losers, I was expecting to see them in Figure 3B, but not all of them were there. Or if it did appear, it was only at one site, but it was mentioned as being an overall winner or loser.}

{\color{red} RGS: We thank the reviewer for clarifying this point. We have now decluttered Figure 3; and highlight taxa that are now mentioned in the text: the Dinoflagellates Karlodinium, Gymnodinaceae and Alexandrium; the Diatoms Stephanodiscus, Nitzschia, Coscinodiscus and Chaetoceros; and the Haptophytes Emiliaina and Chrysochromulina spp.}

@@

\item{Line 122-123: According to Figure 3B, Thalassionema, Navicula, and Nitzschia appear to refer specifically to San Juan. If so, please state this.}

{\color{red} RGS: Our effort has been on visualizing species with more pronounced change, not to highlight species that respond in contrasting ways in the San Juan Islands and the Hood Canal. Most taxa react in the same direction in both regions; with a few exceptions in which species are almost absent on one study area. }

@@

\item{Line 123: Coscinodiscus and Ditylum seem to refer to Hood Canal. This ought to specified as before.}

{\color{red} RGS: Coscinodiscus sees its probability of presence reduced on both Areas. The same happens for D. brightwelli, although our approach for a more readable figure has seen this taxa drop into the Supplementary figures}

@@

\item{Line 129: Write out Emiliania. The ?sp.? after Alexandrium should not be italicized. This sentence might read better if written: ?The coccolithophore Emilinania huxleyi and the dinoflagellate Alexandrium sp. both find more suitable habitat? (rather than ?each finds?)}

{\color{red} RGS: We have updated that sentence, which now reads: "The haptophytes \textit{Emiliania huxleyi} and \textit{Chrysochromulina} sp. the dinoflagellate \textit{Alexandrium} sp. all find more suitable habitat in both of our study areas." Thanks for the suggestion.  }

@@

\item{Line 136-137: This is minor, but this section (or methods) might benefit from a brief mention of why ANOVA was used in one case and Wilcoxon in another.}

{\color{red} RGS: We thank the reviewer for pointing this out: We have rethought this and applied Wilcoxon tests for both global and phylum-specific changes in diversity. As we are comparing pairs of data we prefer not to assume normal distributions of the datasets or the differences between them, Wilcoxon test was a better choice. }

@@

\item{Line 138-141: It seems like this paragraph could be moved to the discussion section.}

{\color{red} RYAN: It is now in the methods, but it has a Discussion feel to it }

@@

\item{Line 147-148: Need a citation for Coscinodiscus as key fish food.}

{\color{red} RGS: Agreed: We now cite the work of J. Zamon (2002) and S. Pasquaud and colleagues (2010) to support the role of Coscinodiscus in the food chain }

### 2020-09-09 Manuscript/What.lies.ahead.reviewed.tex
Updated some text figures and response to reviewee

Given the sea-surface temperatures and pH values for 2017 (observed) and 2095 (estimated) and taxon-specific logistic regression models, we then evaluated the suitability of habitat for each taxon in the future scenario. For each point in the pH - temperature grid, we calculated a species' probability of presence as the mean of 100 independent draws of the posterior model response. For each point, the sum of mean probabilities across species provided richness estimations. The mean value across the 100 draws was the input for a Wilcoxon test of differences in species richness between the 2017 and 2095 scenarios in each region. We performed the Wilcoxon test globally (total species richness) and on each phylum independently.

Taxa surveyed are a function of our metabarcoding PCR primers \cite{leray_new_2013} and reflect the current status of genetic databases, rather than a complete sampling of the planktonic community; we view these results as a cross-section of common taxa useful for understanding the biological effects of ocean conditions.

@@

These communities were distinguished by changes in the relative abundances of a wide variety of taxa. In the Hood Canal for example, the colder water and higher-pH community (community 1 in Fig. 2C) showed higher eDNA indexes of diatoms \textit{Thalassiosira nordenskioeldii}, \textit{Nitzschia cf. promare}, and \textit{Ditylum brightwellii} -- among other species -- relative to a higher-temperature community (community 3 in Fig. 2C). Communities 2 and 3 had relatively higher values for taxa from a mix of groups including \textit{Emiliana huxleyi}, \textit{Minutocellus polymorphus}, \textit{Micromonas pusilla}, and \textit{Bathycoccus prasinos}. Community 3 (Fig. 2A,B) and similar planktonic communities occupy the spectrum of environmental conditions most likely to be encountered in near-future climate scenarios as temperature rises and pH falls. For example, we expect the Hood Canal in 2095 to have the conditions in which community 3 is the most likely community 75\% of the time, an increase of 22\% compared to 2017(see Supplemental Material).

@@

\begin{figure}%[tbhp]
\centering
\includegraphics[width=1\linewidth]{	Figures_for_MS/Figure2.2020-09-03.png}
\caption{\footnotesize{Biological communities and their relationship with environmental variables in the summer months of the Hood Canal. \textbf{A} Constrained Analysis of Principal Coordinates (CAP) of Bray-Curtis dissimilarities among biological communities, as constrained by pH, temperature and salinity (arrows). \textbf{B} Most-likely cluster as a function of temperature and pH given a multinomial logistic model (salinity was uninformative, being tightly correlated with temperature). \textbf{C} Relative abundance (eDNA index; see \cite{kelly2019understanding}) of the taxa best distinguishing the three communities illustrated (SIMPER analysis). See Supporting Information for full analysis.}}
\label{fig:Fig2}
\end{figure}

@@

Likely winners under future conditions are more widely scattered among higher taxonomic groups. The haptophytes \textit{E. huxleyi} and \textit{Chrysochromulina} sp. the dinoflagellate \textit{Alexandrium sp.} each finds more suitable habitat in both of our study areas. Among others not shown in Fig 3, \textit{Chaetoceros} (diatom) and many hydrozoans (Cnidaria) likely increase in San Juan Island, and the potentially fish-killing heterokont flagellate \textit{Pseudochattonella} increases in both study areas. See Supporting Information for a complete list of taxon-specific projections.

@@

%%%%%
 \begin{figure*}
 \centering
 \includegraphics[width=1\linewidth]{Figures_for_MS/Top_fig3_2020-09-03.Other.sel.png}
\caption{\footnotesize{Forecasted changes in plankton in the Salish Sea for Dinoflagellates (panel \textbf{A}), Diatoms (\textbf{B}) and Haptophytes (\textbf{C}). Each panel shows:  (\textbf{1}) probability densities for the occurrence of selected taxa (species- and genus-level) for 2017 (blue) and 2095 (red); data are mean probabilities over 100 model draws, and variance in probability is due to differences in underlying environmental conditions.  (\textbf{2}) Changes in relative species richness within each phylum across the simulated scenario. (\textbf{3}) Relative taxon richness (raster color, warmer colors are more taxon-rich) for each of these same higher-taxa, for plausible ranges of pH and Temperature. Envelopes of observed (2017, blue) and modeled (2095, red) annual conditions in the Salish Sea shown for reference. Hood Canal and San Juan Island plotted separately to illustrate environmental differences between them.}}

@@

Temperate surface oceans worldwide average approximately 14$\degree$C and pH of 8.1 \cite{bopp2013multiple}, and will change substantially in this century \citep[mean $\Delta$ T 2.5$\degree$C, $\Delta$ pH -0.35 globally; RCP 8.5;][]{gattuso2015contrasting}.  Here we document communities exposed to this same range of projected conditions in the present day, along an environmental gradient only ca. 200km wide, allowing us to project future ocean communities from a robust set of underlying observations. Our results reflect patterns in a diverse selection of species from nearshore marine communities in the Salish Sea, consisting of 222 planktonic taxa obtained from the metabarcoding analysis of 227 discrete samples across 77 space-time points (eight sites, 1.5 years). We find that changes in the composition of biological communities closely mirrored the variation in pH and temperature, with clear winners (e.g., \textit{Emiliana huxleyi}, \textit{Alexandrium}, and others) and losers (many, but not all, diatoms) likely to shift the structure and function of future marine communities.

>>>>

Given the sea-surface temperatures and pH values for 2017 (observed) and 2095 (estimated) and taxon-specific logistic regression models, we then evaluated the suitability of habitat for each taxon in the future scenario. For each point in the pH - temperature grid, we calculated a species' probability of presence as the mean of 100 independent draws of the posterior model response. For each point, the sum of mean probabilities across species provided richness estimations. The mean value across the 100 draws was the input for a Wilcoxon test of differences in species richness between the 2017 and 2095 scenarios in each region. We performed the Wilcoxon test globally (total species richness) and on each phylum independently. 
% Reviewer suggests that the paragraoh below belongs to the discussion
{\color{red} RYAN: Move paragraph below to discussion?}
Taxa surveyed are a function of our metabarcoding PCR primers \cite{leray_new_2013} and reflect the current status of genetic databases, rather than a complete sampling of the planktonic community; we view these results as a cross-section of common taxa useful for understanding the biological effects of ocean conditions.

@@

These communities were distinguished by changes in the relative abundances of a wide variety of taxa. In the Hood Canal for example, the colder water and higher-pH community (community 1 in Fig. 2C) showed higher eDNA indexes of diatoms \textit{Thalassiosira nordenskioeldii}, \textit{Nitzschia cf. promare}, and \textit{Ditylum brightwellii} -- among other species -- relative to a higher-temperature community (community 3 in Fig. 2C). Communities 2 and 3 had relatively higher values for taxa from a mix of groups including \textit{Emiliania huxleyi}, \textit{Minutocellus polymorphus}, \textit{Micromonas pusilla}, and \textit{Bathycoccus prasinos}. Community 3 (Fig. 2A,B) and similar planktonic communities occupy the spectrum of environmental conditions most likely to be encountered in near-future climate scenarios as temperature rises and pH falls. For example, we expect the Hood Canal in 2095 to have the conditions in which community 3 is the most likely community 75\% of the time, an increase of 22\% compared to 2017(see Supplemental Material).

@@

\begin{figure}%[tbhp]
\centering
\includegraphics[width=1\linewidth]{	Figures_for_MS/Fig2.png}
\caption{\footnotesize{Biological communities and their relationship with environmental variables in the summer months of the Hood Canal. \textbf{A} Constrained Analysis of Principal Coordinates (CAP) of Bray-Curtis dissimilarities among biological communities, as constrained by pH, temperature and salinity (arrows). \textbf{B} Most-likely cluster as a function of temperature and pH given a multinomial logistic model (salinity was uninformative, being tightly correlated with temperature). \textbf{C} Relative abundance (eDNA index; see \cite{kelly2019understanding}) of the taxa best distinguishing the three communities illustrated (SIMPER analysis). See Supporting Information for full analysis.}}
\label{fig:Fig2}
\end{figure}

@@

Likely winners under future conditions are more widely scattered among higher taxonomic groups. The haptophytes \textit{Emiliania huxleyi} and \textit{Chrysochromulina} sp. the dinoflagellate \textit{Alexandrium} sp. all find more suitable habitat in both of our study areas. Among others not shown in Fig 3, \textit{Chaetoceros} (diatom) and many hydrozoans (Cnidaria) likely increase in San Juan Island, and the potentially fish-killing heterokont flagellate \textit{Pseudochattonella} increases in both study areas. See Supporting Information for a complete list of taxon-specific projections.

@@

%%%%%
 \begin{figure*}
 \centering
 \includegraphics[width=1\linewidth]{Figures_for_MS/Fig3.png}
\caption{\footnotesize{Forecasted changes in plankton in the Salish Sea for Dinoflagellates (panel \textbf{A}), Diatoms (\textbf{B}) and Haptophytes (\textbf{C}). Each panel shows:  (\textbf{1}) probability densities for the occurrence of selected taxa (species- and genus-level) for 2017 (blue) and 2095 (red); data are mean probabilities over 100 model draws, and variance in probability is due to differences in underlying environmental conditions.  (\textbf{2}) Changes in relative species richness within each phylum across the simulated scenario. (\textbf{3}) Relative taxon richness (raster color, warmer colors are more taxon-rich) for each of these same higher-taxa, for plausible ranges of pH and Temperature. Envelopes of observed (2017, blue) and modeled (2095, red) annual conditions in the Salish Sea shown for reference. Hood Canal and San Juan Island plotted separately to illustrate environmental differences between them.}}

@@

Temperate surface oceans worldwide average approximately 14$\degree$C and pH of 8.1 \cite{bopp2013multiple}, and will change substantially in this century \citep[mean $\Delta$ T 2.5$\degree$C, $\Delta$ pH -0.35 globally; RCP 8.5;][]{gattuso2015contrasting}.  Here we document communities exposed to this same range of projected conditions in the present day, along an environmental gradient only ca. 200km wide, allowing us to project future ocean communities from a robust set of underlying observations. Our results reflect patterns in a diverse selection of species from nearshore marine communities in the Salish Sea, consisting of 222 planktonic taxa obtained from the metabarcoding analysis of 227 discrete samples across 77 space-time points (eight sites, 1.5 years). We find that changes in the composition of biological communities closely mirrored the variation in pH and temperature, with clear winners (e.g., \textit{Emiliania huxleyi}, \textit{Alexandrium}, and others) and losers (many, but not all, diatoms) likely to shift the structure and function of future marine communities.
